Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-20 surfaces-write-up/surface-write-up.tex

\DeclareMathOperator{\fr}{frac}
\DeclareMathOperator{\supp}{Supp}
\DeclareMathOperator{\initial}{in_\prec}
\DeclareMathOperator{\gin}{gin}
\DeclareMathOperator{\Eff}{Eff}
\DeclareMathOperator{\sat}{sat}
\DeclareMathOperator{\newspan}{span}
\DeclareMathOperator{\proj}{Proj}
\DeclareMathOperator{\spec}{Spec}
\DeclareMathOperator{\Te}{T_=}
\DeclareMathOperator{\Tp}{T_+}
\DeclareMathOperator{\Tm}{T_-}
\DeclareMathOperator{\num}{p}
\DeclareMathOperator{\den}{q}
\DeclareMathOperator{\lcm}{lcm}
\DeclareMathOperator{\Pic}{Pic}
%\captionsetup[table]{belowskip = 4pt}

\makeatletter
\newcommand{\customlabel}[2]{%
   \protected@write \@auxout {}{\string \newlabel {#1}{{#2}{\thepage}{#2}{#1}{}} }%
   \hypertarget{#1}{#2}
}
\makeatother

@@

\begin{lem}
\label{lem:composite-map}
Let $X$ be an arbitrary scheme and let $D = \sum_{i=0}^{n}\alpha_i D_i\in \bq \otimes_\bz \di X$ where $D_i = V(f_i)$. Suppose we have a
surjection $\phi\colon \bk[w_1,\ldots, w_r] \rightarrow R_D$ given by $w_i
\mapsto p_i(f_0, \ldots, f_n),$ where $p_i$ is a monomial in $f_0,\ldots, f_n$.
Let $\bida_0, \ldots, \bida_n, \bidb_0, \ldots, \bidb_n \in \bz_{\geq 0}.$
Then, define
\begin{align*}
	\Sigma = \langle u^d z_0^{c_0} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=0}^{n} \bida_i c_i = \sum_{i=0}^{n} \bidb_i c_i = 0 \rangle. 
\end{align*}
In this case, we can factor $\phi$ as a composition of $\chi$ and $\psi$ defined by
\[
\begin{tikzcd}
\bk[w_1,\ldots, w_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
w_i \arrow[mapsto]{r} & u^{d_i}z_0^{c_{i0}} \cdots z_n^{c_{in}} \arrow[mapsto]{r} & u^{d_i}f_0^{c_{i0}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$
is at most the maximum of the minimal degree of generation of $\ker \chi$ and
the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
First, note that surjectivity of $\chi$ implies we have an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r}{\rho} & \ker \phi \ar {r}{\tau} & \ker \psi \ar r & 0.
\end{tikzcd}
\]
This shows that lifts of generators of $\ker \psi$ together with images of
generators of $\ker \chi$ generate all of $\ker \phi,$ as desired. 
\end{proof}
\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has
extremal rays $e_0,\ldots, e_t$ in degrees $d_0, \ldots, d_t$ then $\ker \chi$
is generated in degrees at most $2(\sum_{i=0}^{t}d_i-1)$.
\end{lem}
\begin{proof}
Since $e_0, \ldots, e_t$ are extremal rays, Proposition
~\ref{prop:cone-generation} implies every element $\sigma \in \Sigma$ can be written
in a canonical form $\lambda + \sum_{i=0}^{t} \zeta_i e_i$ where all $\zeta_i \in \mathbb{Z}_{\ge 0}$.  Let
$\lambda_0 :=
0,\lambda_1, \ldots, \lambda_r$ be all elements of $\Sigma$ which
can be
written in the form $\lambda_j = \sum_{i=0}^{t}s_i e_i$ with $0
\leq s_i < 1.$ Then, for any $1 \leq j \leq k \leq r,$ we can write $\lambda_j + \lambda_k$ in
the above canonical form, yielding a (possibly trivial) relation in degree at most $\deg \lambda_j
+ \deg \lambda_k\leq 2 \cdot \left( \sum_{i=0}^{t}d_i -1 \right).$
Furthermore, these relations generate all relations, as one can apply a
sequence of these relations to put any $\sigma \in \Sigma$ into canonical form
$\sigma = \lambda + \sum_{i=0}^{t}\zeta_i e_i$ from Proposition
~\ref{prop:cone-generation}.
%as in Equation
%~\eqref{eqn:sigma-canonical-form}.
\end{proof}
\begin{lem}
\label{lem:proj-relations-psi}
Let $D = \sum_{i=0}^{n} \alpha_i D_i \in \bq \otimes_\bz \di \bp^m,$ where
$\deg D_i = \bida_i$, $\alpha_i = \frac{c_i}{k_i}\in \bq$, and
$\ell_i = \lcm_{j \neq i} (k_j)$. Define $\Sigma$ as in Equation
~\eqref{eqn:Sigma-defn} and $\psi$ as ~\eqref{eqn:factor-through-semigroup-ring}. Then, $\ker \psi$ is generated in degrees at most
\begin{align}
\label{eqn:proj-relation-degree}
	\frac{\max_{0\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=0}^n \ell_i a_i.
\end{align}

@@

\section{Acknowledgments}
\label{sec:ack}
We are grateful to David Zureick-Brown for suggesting this problem 
and for providing invaluable guidance. We thank Shou-Wu Zhang and Peter
Landesman for reading this paper and proving helpful suggestions.
We also thank Ken Ono and the
Emory University Number Theory REU for arranging our project and
providing a great environment for mathematical learning and
collaboration.
Finally, we gratefully acknowledge the financial support given by
NSF Grant Award Number 1250467 via the Emory University Number
Theory REU. We deeply appreciate all of the support that has made
our work possible.

>>>>

\DeclareMathOperator{\fr}{frac}
\DeclareMathOperator{\supp}{Supp}
\DeclareMathOperator{\initial}{in_\prec}
\DeclareMathOperator{\gin}{gin}
\DeclareMathOperator{\Eff}{Eff}
\DeclareMathOperator{\sat}{sat}
\DeclareMathOperator{\newspan}{span}
\DeclareMathOperator{\proj}{Proj}
\DeclareMathOperator{\spec}{Spec}
\DeclareMathOperator{\Te}{T_=}
\DeclareMathOperator{\Tp}{T_+}
\DeclareMathOperator{\Tm}{T_-}
\DeclareMathOperator{\num}{p}
\DeclareMathOperator{\den}{q}
\DeclareMathOperator{\lcm}{lcm}
\DeclareMathOperator{\Pic}{Pic}
\makeatletter
\newcommand{\customlabel}[2]{%
   \protected@write \@auxout {}{\string \newlabel {#1}{{#2}{\thepage}{#2}{#1}{}} }%
   \hypertarget{#1}{#2}
}
\makeatother

@@

\begin{lem}
\label{lem:composite-map}
Let $X$ be an arbitrary scheme and let $D = \sum_{i=0}^{n}\alpha_i D_i\in \bq \otimes_\bz \di X$ where $D_i = V(f_i)$. Suppose we have a
surjection $\phi\colon \bk[w_1,\ldots, w_r] \rightarrow R_D$ given by $w_i
\mapsto p_i(f_0, \ldots, f_n),$ where $p_i$ is a monomial in $f_0,\ldots, f_n$.
Let $\bida_0, \ldots, \bida_n, \bidb_0, \ldots, \bidb_n \in \bz_{\geq 0}.$
Then, define
\begin{align*}
	\Sigma = \langle u^d z_0^{c_0} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=0}^{n} \bida_i c_i = \sum_{i=0}^{n} \bidb_i c_i = 0 \rangle. 
\end{align*}
In this case, we can factor $\phi$ as a composition of $\chi$ and $\psi$ defined by
\[
\begin{tikzcd}
\bk[w_1,\ldots, w_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
w_i \arrow[mapsto]{r} & u^{d_i}z_0^{c_{i0}} \cdots z_n^{c_{in}} \arrow[mapsto]{r} & u^{d_i}f_0^{c_{i0}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$
is at most the maximum of the minimal degree of generation of $\ker \chi$ and
the minimal degree of generation of $\ker \psi$.
\end{lem}

\begin{proof}
First, note that surjectivity of $\chi$ implies we have an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r}{\rho} & \ker \phi \ar {r}{\tau} & \ker \psi \ar r & 0.
\end{tikzcd}
\]

This shows that lifts of generators of $\ker \psi$ together with images of
generators of $\ker \chi$ generate all of $\ker \phi,$ as desired. 
\end{proof}

\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has
extremal rays $e_0,\ldots, e_t$ in degrees $d_0, \ldots, d_t$ then $\ker \chi$
is generated in degrees at most $2(\sum_{i=0}^{t}d_i-1)$.
\end{lem}
\begin{proof}
Since $e_0, \ldots, e_t$ are extremal rays, Proposition
~\ref{prop:cone-generation} implies every element $\sigma \in \Sigma$ can be written
in a canonical form $\lambda + \sum_{i=0}^{t} \zeta_i e_i$ where all $\zeta_i \in \mathbb{Z}_{\ge 0}$.  Let
$\lambda_0 :=
0,\lambda_1, \ldots, \lambda_r$ be all elements of $\Sigma$ which
can be
written in the form $\lambda_j = \sum_{i=0}^{t}s_i e_i$ with $0
\leq s_i < 1.$ Then, for any $1 \leq j \leq k \leq r,$ we can write $\lambda_j + \lambda_k$ in
the above canonical form, yielding a (possibly trivial) relation in degree at most $\deg \lambda_j
+ \deg \lambda_k\leq 2 \cdot \left( \sum_{i=0}^{t}d_i -1 \right).$
Furthermore, these relations generate all relations, as one can apply a
sequence of these relations to put any $\sigma \in \Sigma$ into canonical form
$\sigma = \lambda + \sum_{i=0}^{t}\zeta_i e_i$ from Proposition
~\ref{prop:cone-generation}.
\end{proof}
\begin{lem}
\label{lem:proj-relations-psi}
Let $D = \sum_{i=0}^{n} \alpha_i D_i \in \bq \otimes_\bz \di \bp^m,$ where
$\deg D_i = \bida_i$, $\alpha_i = \frac{c_i}{k_i}\in \bq$, and
$\ell_i = \lcm_{j \neq i} (k_j)$. Define $\Sigma$ as in Equation
~\eqref{eqn:Sigma-defn} and $\psi$ as ~\eqref{eqn:factor-through-semigroup-ring}. Then, $\ker \psi$ is generated in degrees at most
\begin{align}
\label{eqn:proj-relation-degree}
	\frac{\max_{0\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=0}^n \ell_i a_i.
\end{align}

@@

\section{Acknowledgments}
\label{sec:ack}
We are grateful to David Zureick-Brown for offering great advice on
this project, for many useful suggestions on this paper, and for
many valuable conversations. We thank Peter
Landesman and Shou-Wu Zhang for offering helpful comments and
suggestions for this paper. We also thank Ken Ono and the
Emory University Number Theory REU for arranging our project and
for the great support that we have received. Finally, we would like to acknowledge the
support of the National Science Foundation (grant number DMS-1250467
).
